The formal verification of the ctm approach to forcing \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}

### 2022-08-11 Independence_CH/independence_ch_isabelle.tex
Target journal now unspecified
One bib, one more entry
Another possible title
typo :facepalm:
First brief description of the discipline.
ITP -> assistant. Updated abstract

\begin{document}
%
\title{Some lessons after the formalization of the ctm approach to forcing%
  \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
}
%
\titlerunning{Lessons after formalizing ctm forcing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Emmanuel Gunther\inst{1} \and
Miguel Pagano\inst{1} \and \\
Pedro Sánchez Terraf\inst{1,2}%\orcidID{0000-0003-3928-6942}
\and
Matías Steinberg\inst{1}
}
%
\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf, M.~Steinberg}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidad Nacional de C\'ordoba. 
  \\  Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
  Computaci\'on.
  \and
    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF),
    Conicet. C\'ordoba. Argentina. \\
    \email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar\\
        matias.steinberg@mi.unc.edu.ar}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  We'll discuss some highlights of our computer-verified
  proof of the construction, given a countable transitive set model $M$
  of $\ZFC$, of generic extensions  satisfying $\ZFC+\neg\CH$ and $\ZFC+\CH$.
  In particular,
  we isolated a set $\Delta$ of 39 instances
  of Axiom of Replacement and a function $F$
  such that such that for any finite fragment $\Phi\sbq\ZFC$,
  $F(\Phi)\sbq\ZFC$ is also finite and if
  $M\models F(\Phi) + \Delta$ then $M[G]\models \Phi + \neg \CH$.
  We also obtained the formulas yielded by the Forcing Definability Theorem
  explicitly.

@@

  We'll also compare our formalization with the recent one by Jesse
  M.~Han and Floris van Doorn in the proof assistant \emph{Lean}.

@@

\input{details}
%% 
%% \input{forcing_theorems}
%% 
%% \input{proper_extension}
%% 
%% \input{replacement_choice}
%% 
\input{random}

@@

\input{conclusion}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
%\bibliographystyle{splncs04}
\bibliographystyle{mi-estilo-else}
\bibliography{independence_ch_isabelle,clam2021}

\end{document}

>>>>

\begin{document}
%
\title{The formal verification of the ctm approach to forcing%Some lessons after the formalization of the ctm approach to forcing%
  \thanks{Supported by Secyt-UNC project 33620180100465CB and Conicet.}%
}
%
\titlerunning{Formalization of ctm forcing}%Lessons after formalizing ctm forcing}
% If the paper title is too long for the running head, you can set
% an abbreviated paper title here
%
\author{Emmanuel Gunther\inst{1} \and
Miguel Pagano\inst{1} \and \\
Pedro Sánchez Terraf\inst{1,2}%\orcidID{0000-0003-3928-6942}
\and
Matías Steinberg\inst{1}
}
%
\authorrunning{E.~Gunther, M.~Pagano, P.~Sánchez Terraf, M.~Steinberg}
% First names are abbreviated in the running head.
% If there are more than two authors, 'et al.' is used.
%
\institute{Universidad Nacional de C\'ordoba. 
  \\  Facultad de Matem\'atica, Astronom\'{\i}a,  F\'{\i}sica y
  Computaci\'on.
  \and
    Centro de Investigaci\'on y Estudios de Matem\'atica (CIEM-FaMAF),
    Conicet. C\'ordoba. Argentina. \\
    \email{\{gunther,pagano,sterraf\}@famaf.unc.edu.ar\\
        matias.steinberg@mi.unc.edu.ar}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
  We discuss some highlights of our computer-verified
  proof of the construction, given a countable transitive set-model $M$
  of $\ZFC$, of generic extensions  satisfying $\ZFC+\neg\CH$ and $\ZFC+\CH$.
  Moreover, let $\calR$ be the set of instances of the Axiom of
  Replacement. We isolated a 32-element subset $\Omega\sbq\calR$ and
  defined $\calF:\calR\to\calR$
  such that such that for every finite $\Phi\sbq\calR$
  we have
  $M\models Z \cup \calF\dimg\Phi \cup \Delta$ implies $M[G]\models Z
  \cup \Phi \cup \{ \neg \CH \}$ for any $M$-generic $G$, where $Z$ is
  Zermelo set theory.
  We also obtained the formulas yielded by the Forcing Definability Theorem
  explicitly.

@@

  We'll also compare our formalization with the recent one by J.~M.~Han and F.~van Doorn in the proof assistant \emph{Lean}.

@@

\input{details}
%% 
%% \input{forcing_theorems}
%% 
%% \input{proper_extension}
%% 
%% \input{replacement_choice}
%%

@@

\input{conclusion}
%
% ---- Bibliography ----
%
% BibTeX users should specify bibliography style 'splncs04'.
% References will then be sorted and formatted in the correct style.
%
%\bibliographystyle{splncs04}
\bibliographystyle{mi-estilo-else}
\bibliography{independence_ch_isabelle}

\input{discipline}
\end{document}

### 2022-08-11 Independence_CH/intro.tex
ITP -> assistant. Updated abstract

\subsection{Our achievements}
We formalized a model-theoretic rendition of forcing (Section~\ref{sec:forcing}), showing how to
construct proper extension of ctms of $\ZF$ (respectively, with
$\AC$), and we formalized the basic forcing notions required to obtain
ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$. No meta-theoretic issues
(consistency, FOL calculi, etc) were formalized, so we were merely
concerned with the mathematics of forcing. Nevertheless, by inspecting
the foundations underlying our proof assistant Isabelle
(Section~\ref{sec:isabelle-metalogic-meta}) it can be stated that our
formalization is closest to a proof in $\ZF$ of the previous
implications.

>>>>

\subsection{Our achievements}
We formalized a model-theoretic rendition of forcing (Section~\ref{sec:forcing}), showing how to
construct proper extensions of ctms of $\ZF$ (respectively, with
$\AC$), and we formalized the basic forcing notions required to obtain
ctms of $\ZFC + \neg\CH$ and of $\ZFC + \CH$. No meta-theoretic issues
(consistency, FOL calculi, etc) were formalized, so we were merely
concerned with the mathematics of forcing. Nevertheless, by inspecting
the foundations underlying our proof assistant Isabelle
(Section~\ref{sec:isabelle-metalogic-meta}) it can be stated that our
formalization is closest to a proof in $\ZF$ of the previous
constructions.

### 2022-08-11 Independence_CH/proof_assists_isabelle.tex
ITP -> assistant. Updated abstract

Let us briefly introduce Isabelle in the large landscape of proof
assistans (or ITPs for interactive theorem provers); we refer to the
excelent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
thorough reconstruction of the history of ITPs.

It is expected that an ITP assists the human user while mechanizing
some piece of mathematics; the interaction varies from system to
system, but a common interface consists on the system showing the
current goal and context. The user instructs the ITP to modify the
goal/context by means of tactics; a forward rule changes the context
with the intention of having in the new context a hypothesis closer to
the goal. A backward rule represents a deduction whose conclusion is
the goal, thus the new goal corresponds to the premises of the rule. A
proof is completed when the (current) goal is an instance of a
hypothesis (in the current context).

@@

To have any value at all, the system should be able to say if a tactic
makes sense (for instance, it must be forbidden to use the backward
tactic \textit{conjI} when the main connective is a
disjunction). Moreover one should be able to ascertain the validity of
the reaction of an ITP (be its rejection or its acceptance of some
alleged logical step). Edinburgh LCF~\cite{Gordon1979-qm} was an
influential proof assistant where that verification could be achieved
by checking a small \emph{kernel}; this is the only module that
constructs proofs in response to user scripts. The rest of the system
can interact with the kernel via an interface that exposes functions
guaranteeing the validity of the proofs.

@@

In contrast with LCF, some of the more prominent ITPs of today are
based on some type theory. Both Coq~\cite{coq} and
lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
Inductive Constructions
\cite{DBLP:conf/mfps/PfenningP89,DBLP:journals/iandc/CoquandH88},
while Agda~\cite{agda} is an extension of Martin-Löf type
theory~\cite{DBLP:books/daglib/0000395}. Mizar~\cite{mizar} is the
oldest ITP still used today but is far away both in terms of
foundations and architecture from Isabelle; Mizar inspired, though, the
Isar dialect used in nowadays Isabelle.

>>>>

Let us briefly introduce Isabelle in the large landscape of \emph{proof
assistants} (“assitants” for short; also known as “interactive theorem provers”); we refer to the
excellent chapter by \citet{DBLP:series/hhl/HarrisonUW14} for a more
thorough reconstruction of the history of assistants.

It is expected that an assistant aids the human user while mechanizing
some piece of mathematics; the interaction varies from system to
system, but a common interface consists on the system showing the
current goal and context. The user instructs the assistant to modify the
goal/context by means of tactics; a forward rule changes the context
with the intention of having in the new context a hypothesis closer to
the goal. A backward rule represents a deduction whose conclusion is
the goal, thus the new goal corresponds to the premises of the rule. A
proof is completed when the (current) goal is an instance of a
hypothesis (in the current context).

@@

To have any value at all, the system should be able to say if a tactic
makes sense (for instance, it must be forbidden to use the backward
tactic \textit{conjI} when the main connective is a
disjunction). Moreover one should be able to ascertain the validity of
the reaction of an assistant (be its rejection or its acceptance of some
alleged logical step). Edinburgh LCF~\cite{Gordon1979-qm} was an
influential proof assistant where that verification could be achieved
by checking a small \emph{kernel}; this is the only module that
constructs proofs in response to user scripts. The rest of the system
can interact with the kernel via an interface that exposes functions
guaranteeing the validity of the proofs.

@@

In contrast with LCF, some of the more prominent assistants of today are
based on some type theory. Both Coq~\cite{coq} and
Lean~\cite{DBLP:conf/cade/Moura021} are based on the Calculus of
Inductive Constructions
\cite{DBLP:conf/mfps/PfenningP89,DBLP:journals/iandc/CoquandH88},
while Agda~\cite{agda} is an extension of Martin-Löf type
theory~\cite{DBLP:books/daglib/0000395}. Mizar~\cite{mizar} is the
oldest assistant still used today but is far away both in terms of
foundations and architecture from Isabelle; Mizar inspired, though, the
Isar dialect used in Isabelle nowadays.
